---
title: "Bayesian"
author: "Qingyang Du"
date: "2026-02-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###############################################################################
#                      BAYESIAN ANALYSIS SCRIPT
#
# Project:  Treatment Response to HAART and Hard Drug Use
# Prereq:   Run 01_data_cleaning_v3.R (produces hiv_cleaned.RData with dat_cc)
#
# Approach: Follows the professor's cmdstanr workflow:
#   1. Write a generic Stan linear regression model
#   2. Compile once, reuse for all outcomes
#   3. Prepare data (design matrix + priors) per outcome
#   4. Fit via MCMC (4 chains)
#   5. Summarize posteriors (mean, SD, MCSE, 95% HPDI, ESS)
#   6. Model fit statistics (WAIC, LOO-IC, DIC)
#   7. MCMC diagnostics (trace, density overlay, ACF, R-hat)
#
# Models (for each of 4 outcomes):
#   Model 1: Adjusted — Y2 ~ hard_drugs_bl + Y_bl + confounders
#   Model 2: Adjusted + Adherence — Model 1 + ADH_cat
###############################################################################


# =============================================================================
# 0.  LOAD PACKAGES AND DATA
# =============================================================================
```{r}
# install.packages("cmdstanr", repos = c('https://stan-dev.r-universe.dev',
#                                         getOption("repos")))
# library(cmdstanr)
# install_cmdstan()   # Run once to install CmdStan backend

library(cmdstanr)
library(bayesplot)   # MCMC diagnostic plots
library(posterior)   # summarizing posterior draws
library(bayestestR)  # highest density posterior intervals (hdi)
library(mcmcse)      # Monte Carlo standard errors
library(loo)         # WAIC and LOO-IC
library(dplyr)
library(tibble)

load("hiv_cleaned.RData")   # contains dat_cc




# =============================================================================
# 1.  WRITE THE STAN MODEL (reusable for all outcomes)
# =============================================================================
# Same structure as the professor's example: normal priors on beta,
# half-normal prior on sigma, generated quantities for log-likelihood.

stan_code <- "data {
  int<lower=0> N;                  // number of observations
  int<lower=0> P;                  // number of predictors including intercept
  matrix[N, P] X;                  // design matrix (first column = intercept)
  vector[N] y;                     // outcome

  vector[P] prior_mean;            // prior means for each beta
  vector<lower=0>[P] prior_sd;     // prior SDs for each beta

  real<lower=0> sigma_prior_sd;    // SD for half-normal prior on sigma
}

parameters {
  vector[P] beta;                  // regression coefficients
  real<lower=0> sigma;             // residual SD
}

model {
  // Priors for regression coefficients
  beta ~ normal(prior_mean, prior_sd);

  // Half-normal prior for sigma
  sigma ~ normal(0, sigma_prior_sd);

  // Likelihood
  y ~ normal(X * beta, sigma);
}

generated quantities {
  // Log likelihood for each observation (for WAIC / LOO-IC)
  vector[N] log_lik;
  for (n in 1:N) {
    log_lik[n] = normal_lpdf(y[n] | X[n] * beta, sigma);
  }
}
"

stan_file <- write_stan_file(stan_code, dir = ".", basename = "linear_regression")


# =============================================================================
# 2.  COMPILE THE STAN MODEL
# =============================================================================

mod <- cmdstan_model(stan_file)


# =============================================================================
# 3.  DEFINE OUTCOMES AND PRIOR SPECIFICATIONS
# =============================================================================

# Outcome definitions — same structure as frequentist script
outcomes <- list(
  list(name  = "delta_log_VLOAD",
       y     = "delta_log_VLOAD",
       label = "Change in Log10 Viral Load"),
  list(name  = "delta_LEU3N",
       y     = "delta_LEU3N",
       label = "Change in CD4+ T Cell Count"),
  list(name  = "delta_AGG_PHYS",
       y     = "delta_AGG_PHYS",
       label = "Change in Physical QoL"),
  list(name  = "delta_AGG_MENT",
       y     = "delta_AGG_MENT",
       label = "Change in Mental QoL")
)

# Confounders (same for all outcomes)
confounders <- c("age_bl", "RACE_bin_bl", "EDUC_bin_bl", "BMI_bl", "SMOKE_bin_bl")


# =============================================================================
# 4.  HELPER FUNCTION: PREPARE DATA AND SET PRIORS
# =============================================================================

prepare_stan_data <- function(dat, outcome_info, include_adherence = FALSE) {

  # Build the model formula
  rhs <- c("hard_drugs_bl", confounders)
  if (include_adherence) rhs <- c(rhs, "ADH_cat")
  f <- as.formula(paste(outcome_info$y, "~", paste(rhs, collapse = " + ")))

  # Design matrix
  X <- model.matrix(f, data = dat)
  y <- dat[[outcome_info$y]]

  N <- nrow(X)
  P <- ncol(X)

  # --- Prior specification ---
  # Weakly informative priors scaled to the outcome:
  #   Intercept: Normal(mean(y), 2 * sd(y))  — centered on the data mean
  #   All other coefficients: Normal(0, sd(y)) — centered at no effect,
  #     scaled to the outcome's SD so they are weakly informative
  #   Sigma: half-Normal(0, 2 * sd(y))

  y_sd <- sd(y)
  y_mean <- mean(y)

  m <- c(y_mean, rep(0, P - 1))       # prior means
  s <- c(2 * y_sd, rep(y_sd, P - 1))  # prior SDs
  sigma_sd <- 2 * y_sd                 # sigma prior SD

  data_list <- list(
    N = N,
    P = P,
    X = X,
    y = y,
    prior_mean = m,
    prior_sd = s,
    sigma_prior_sd = sigma_sd
  )

  list(data_list = data_list, X = X, y = y, P = P, N = N,
       col_names = colnames(X))
}


# =============================================================================
# 5.  HELPER FUNCTION: SUMMARIZE POSTERIOR
# =============================================================================

summarize_posterior <- function(fit, col_names) {
  draws_mat <- as_draws_matrix(fit$draws())
  params <- colnames(draws_mat)
  params <- params[!grepl("lp__|log_lik", params)]

  summary_table <- lapply(params, function(p) {
    vals <- as.numeric(draws_mat[, p])

    # Monte Carlo standard error
    mcse_val <- mcmcse::mcse(vals)$se

    # Effective sample size
    ess_val <- ess_bulk(vals)

    # 95% Highest Posterior Density Interval
    hpd <- hdi(vals, ci = 0.95)

    tibble(
      Parameter     = p,
      Estimate      = mean(vals),
      MCSE          = mcse_val,
      Posterior_SD  = sd(vals),
      HPDI_2.5      = hpd$CI_low,
      HPDI_97.5     = hpd$CI_high,
      ESS           = ess_val
    )
  }) %>% bind_rows()

  # Map beta indices to variable names
  beta_idx <- grep("^beta\\[", summary_table$Parameter)
  if (length(beta_idx) == length(col_names)) {
    summary_table$Variable <- NA_character_
    summary_table$Variable[beta_idx] <- col_names
    summary_table$Variable[summary_table$Parameter == "sigma"] <- "sigma"
    summary_table <- summary_table %>%
      select(Parameter, Variable, everything())
  }

  # Check MCSE < 6% of posterior SD
  summary_table <- summary_table %>%
    mutate(MCSE_pct_SD = round(100 * MCSE / Posterior_SD, 2),
           MCSE_OK     = MCSE_pct_SD < 6)

  summary_table
}


# =============================================================================
# 6.  HELPER FUNCTION: MODEL FIT STATISTICS
# =============================================================================

compute_fit_stats <- function(fit, X, y) {
  draws_mat <- as_draws_matrix(fit$draws())
  loglik_mat <- as_draws_matrix(fit$draws("log_lik"))

  # LOO-IC
  loo_res <- loo(loglik_mat)

  # WAIC
  waic_res <- waic(loglik_mat)

  # DIC (computed by hand, following professor's approach)
  D_theta <- -2 * rowSums(loglik_mat)
  mean_D <- mean(D_theta)

  beta_mean <- colMeans(draws_mat[, grep("^beta\\[", colnames(draws_mat))])
  sigma_mean <- mean(draws_mat[, "sigma"])
  D_bar_theta <- -2 * sum(dnorm(y, mean = X %*% beta_mean,
                                 sd = sigma_mean, log = TRUE))
  DIC <- 2 * mean_D - D_bar_theta

  list(loo = loo_res, waic = waic_res, DIC = DIC)
}


# =============================================================================
# 7.  HELPER FUNCTION: MCMC DIAGNOSTICS
# =============================================================================

run_diagnostics <- function(fit, col_names, outcome_label, model_label) {

  draws_array <- as_draws_array(fit$draws())

  # Build parameter names for diagnostic plots
  P <- length(col_names)
  param_names <- c(paste0("beta[", 1:P, "]"), "sigma")

  # Trace plots
  p_trace <- mcmc_trace(draws_array, pars = param_names) +
    ggtitle(paste(outcome_label, "-", model_label, ": Trace Plots"))
  print(p_trace)

  # Posterior density overlay across chains
  p_dens <- mcmc_dens_overlay(draws_array, pars = param_names) +
    ggtitle(paste(outcome_label, "-", model_label, ": Posterior Densities"))
  print(p_dens)

  # Autocorrelation
  p_acf <- mcmc_acf(draws_array, pars = param_names) +
    ggtitle(paste(outcome_label, "-", model_label, ": Autocorrelation"))
  print(p_acf)

  # R-hat and ESS from cmdstanr
  rhat_summary <- fit$summary(variables = param_names)
  print(rhat_summary)

  # CmdStan diagnostics (divergences, treedepth, energy)
  fit$cmdstan_diagnose()
}

```

```{r}
# =============================================================================
# 8.  FIT ALL MODELS
# =============================================================================

bayes_results <- list()

for (o in outcomes) {

  message("\n============================================================")
  message("FITTING: ", o$label)
  message("============================================================\n")

  # --- Model 1: Adjusted (no adherence) ---
  prep1 <- prepare_stan_data(dat_cc, o, include_adherence = FALSE)

  fit1 <- mod$sample(
    data = prep1$data_list,
    chains = 4,
    iter_warmup = 2500,
    iter_sampling = 10000,
    seed = 42,
    parallel_chains = 4
  )

  summ1 <- summarize_posterior(fit1, prep1$col_names)
  fits1 <- compute_fit_stats(fit1, prep1$X, prep1$y)

  message("\n--- Adjusted Model Summary ---")
  print(summ1)
  message("DIC: ", round(fits1$DIC, 2))

  # --- Model 2: Adjusted + Adherence ---
  prep2 <- prepare_stan_data(dat_cc, o, include_adherence = TRUE)

  fit2 <- mod$sample(
    data = prep2$data_list,
    chains = 4,
    iter_warmup = 2500,
    iter_sampling = 10000,
    seed = 42,
    parallel_chains = 4
  )

  summ2 <- summarize_posterior(fit2, prep2$col_names)
  fits2 <- compute_fit_stats(fit2, prep2$X, prep2$y)

  message("\n--- Adjusted + Adherence Model Summary ---")
  print(summ2)
  message("DIC: ", round(fits2$DIC, 2))

  # Store results
  bayes_results[[o$name]] <- list(
    label        = o$label,
    # Adjusted model
    fit_adj      = fit1,
    summary_adj  = summ1,
    fitstats_adj = fits1,
    prep_adj     = prep1,
    # Adjusted + adherence
    fit_adj_adh      = fit2,
    summary_adj_adh  = summ2,
    fitstats_adj_adh = fits2,
    prep_adj_adh     = prep2
  )
}

```

```{r}
# =============================================================================
# 9.  EXTRACT HARD DRUG USE EFFECT ACROSS ALL MODELS
# =============================================================================

extract_drug_bayes <- function(summary_tbl, model_label, outcome_label) {
  summary_tbl %>%
    filter(Variable == "hard_drugs_blYes") %>%
    mutate(Outcome = outcome_label, Model = model_label) %>%
    select(Outcome, Model, Estimate, Posterior_SD, HPDI_2.5, HPDI_97.5,
           ESS, MCSE, MCSE_pct_SD, MCSE_OK)
}

bayes_drug_effects <- map_dfr(names(bayes_results), function(nm) {
  r <- bayes_results[[nm]]
  bind_rows(
    extract_drug_bayes(r$summary_adj,     "Adjusted",             r$label),
    extract_drug_bayes(r$summary_adj_adh, "Adjusted + Adherence", r$label)
  )
})

bayes_drug_effects <- bayes_drug_effects %>%
  mutate(across(c(Estimate, Posterior_SD, HPDI_2.5, HPDI_97.5, MCSE),
                ~ round(., 4)),
         ESS = round(ESS, 0))

bayes_drug_effects


# =============================================================================
# 10.  POSTERIOR PROBABILITY OF CLINICALLY MEANINGFUL DIFFERENCE
# =============================================================================
# Clinically meaningful differences (per investigator):
#   log10(VLOAD): 0.5 log10 copies/mL
#   CD4+ (LEU3N): 50 cells
#   AGG_PHYS:     2 points
#   AGG_MENT:     2 points
#
# For change scores, beta = difference in change between drug users vs non-users.
# Worse outcomes for drug users means:
#   delta_log_VLOAD:  less VL suppression → beta > 0   → clinically meaningful if beta > 0.5
#   delta_LEU3N:      less CD4 recovery   → beta < 0   → clinically meaningful if beta < -50
#   delta_AGG_PHYS:   less QoL gain       → beta < 0   → clinically meaningful if beta < -2
#   delta_AGG_MENT:   less QoL gain       → beta < 0   → clinically meaningful if beta < -2

hypothesis_thresholds <- list(
  "delta_log_VLOAD" = list(direction = "greater", threshold = 0.5,
                           label = "P(beta > 0.5)"),
  "delta_LEU3N"     = list(direction = "less",    threshold = -50,
                           label = "P(beta < -50)"),
  "delta_AGG_PHYS"  = list(direction = "less",    threshold = -2,
                           label = "P(beta < -2)"),
  "delta_AGG_MENT"  = list(direction = "less",    threshold = -2,
                           label = "P(beta < -2)")
)

posterior_probs <- map_dfr(names(bayes_results), function(nm) {
  r <- bayes_results[[nm]]
  h <- hypothesis_thresholds[[nm]]

  # Extract hard_drugs_bl posterior draws from the adjusted model
  draws_mat <- as_draws_matrix(r$fit_adj$draws())
  drug_col <- which(r$prep_adj$col_names == "hard_drugs_blYes")
  drug_draws <- as.numeric(draws_mat[, paste0("beta[", drug_col, "]")])

  # P(effect in hypothesized direction) — any effect
  if (h$direction == "greater") {
    prob_direction <- mean(drug_draws > 0)
  } else {
    prob_direction <- mean(drug_draws < 0)
  }

  # P(effect exceeds clinically meaningful threshold)
  if (h$direction == "greater") {
    prob_clinical <- mean(drug_draws > h$threshold)
  } else {
    prob_clinical <- mean(drug_draws < h$threshold)
  }

  tibble(
    Outcome              = r$label,
    P_direction          = round(prob_direction, 4),
    Clinical_threshold   = h$threshold,
    P_clinically_meaningful = round(prob_clinical, 4),
    Threshold_label      = h$label
  )
})

posterior_probs


# =============================================================================
# 11.  MEDIATION COMPARISON (BAYESIAN)
# =============================================================================

bayes_mediation <- bayes_drug_effects %>%
  select(Outcome, Model, Estimate) %>%
  pivot_wider(names_from = Model, values_from = Estimate,
              names_prefix = "est_") %>%
  rename(est_adj     = `est_Adjusted`,
         est_adj_adh = `est_Adjusted + Adherence`) %>%
  mutate(
    change     = est_adj_adh - est_adj,
    pct_change = round(100 * change / est_adj, 1)
  )

bayes_mediation


# =============================================================================
# 12.  MCMC DIAGNOSTICS — ALL MODELS
# =============================================================================

for (nm in names(bayes_results)) {
  r <- bayes_results[[nm]]

  message("\n--- Diagnostics: ", r$label, " (Adjusted) ---")
  run_diagnostics(r$fit_adj, r$prep_adj$col_names,
                  r$label, "Adjusted")

  message("\n--- Diagnostics: ", r$label, " (Adjusted + Adherence) ---")
  run_diagnostics(r$fit_adj_adh, r$prep_adj_adh$col_names,
                  r$label, "Adjusted + Adherence")
}


# =============================================================================
# 13.  MODEL FIT COMPARISON TABLE
# =============================================================================

fit_comparison <- map_dfr(names(bayes_results), function(nm) {
  r <- bayes_results[[nm]]
  tibble(
    Outcome         = r$label,
    # Adjusted model
    DIC_adj         = round(r$fitstats_adj$DIC, 2),
    WAIC_adj        = round(r$fitstats_adj$waic$estimates["waic", "Estimate"], 2),
    LOOIC_adj       = round(r$fitstats_adj$loo$estimates["looic", "Estimate"], 2),
    # Adjusted + adherence
    DIC_adj_adh     = round(r$fitstats_adj_adh$DIC, 2),
    WAIC_adj_adh    = round(r$fitstats_adj_adh$waic$estimates["waic", "Estimate"], 2),
    LOOIC_adj_adh   = round(r$fitstats_adj_adh$loo$estimates["looic", "Estimate"], 2)
  )
})

fit_comparison



```

```{r}
###############################################################################
#  DIAGNOSTIC FIGURES — Add to end of Bayesian.Rmd
#
#  Figure 1: Posterior density of hard drug use coefficient (4-panel)
#  Figure 2: Frequentist Q-Q plots for residuals (4-panel)
#
#  Prereqs: bayes_results (from Bayesian.Rmd), freq_results.RData
###############################################################################

library(tidyverse)

# =============================================================================
# FIGURE 1: POSTERIOR DENSITY PLOTS
# =============================================================================
# Extract posterior draws for the hard drug use coefficient (beta[1])
# from each outcome's adjusted model fit

outcome_names <- names(bayes_results)

# Clinical thresholds for each outcome (on the change-score scale)
# Viral load: worse = beta > 0, threshold = 0.5
# CD4+:       worse = beta < 0, threshold = -50
# Phys QoL:   worse = beta < 0, threshold = -2
# Ment QoL:   worse = beta < 0, threshold = -2
thresholds <- c(0.5, -50, -2, -2)

# Short labels for panels
panel_labels <- c(
  expression(Delta ~ Log[10] ~ "Viral Load"),
  expression(Delta ~ "CD4+ T Cell Count"),
  expression(Delta ~ "Physical QoL"),
  expression(Delta ~ "Mental QoL")
)

# Extract draws into a single data frame
draws_list <- map2_dfr(outcome_names, seq_along(outcome_names), function(nm, i) {
  fit <- bayes_results[[nm]]$fit_adj
  # beta[1] = hard drug use (first predictor after intercept alpha)
  d <- as.numeric(fit$draws(variables = "beta[2]", format = "matrix"))
  tibble(
    Outcome  = nm,
    panel_id = i,
    draw     = d
  )
})

# Assign panel labels as factor for faceting
draws_list <- draws_list %>%
  mutate(panel_label = factor(panel_id,
                              levels = 1:4,
                              labels = c("Delta Log10 Viral Load",
                                         "Delta CD4+ T Cell Count",
                                         "Delta Physical QoL",
                                         "Delta Mental QoL")))

# Threshold and zero reference lines per panel
ref_lines <- tibble(
  panel_label = factor(c("Delta Log10 Viral Load",
                          "Delta CD4+ T Cell Count",
                          "Delta Physical QoL",
                          "Delta Mental QoL"),
                        levels = levels(draws_list$panel_label)),
  threshold   = thresholds
)

fig1 <- ggplot(draws_list, aes(x = draw)) +
  geom_density(fill = "#3182bd", alpha = 0.4, colour = "#08519c") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_vline(data = ref_lines, aes(xintercept = threshold),
             linetype = "dotted", colour = "firebrick", linewidth = 0.8) +
  facet_wrap(~ panel_label, scales = "free", ncol = 2) +
  labs(
    title    = "Posterior Density of Hard Drug Use Effect (Adjusted Model)",
    subtitle = "Dashed grey = null (0); dotted red = clinical threshold",
    x = "Posterior Estimate",
    y = "Density"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text      = element_text(face = "bold"),
    plot.title       = element_text(size = 12, face = "bold"),
    plot.subtitle    = element_text(size = 9, colour = "grey40"),
    panel.grid.minor = element_blank()
  )

ggsave("fig1_posterior_densities.png", fig1,
       width = 8, height = 6, dpi = 300)

```


```{R}
###############################################################################
#  COMBINED DIAGNOSTIC FIGURES — All Parameters, Adjusted Model
#
#  For each of the 3 diagnostic types (trace, density, ACF), produces
#  one figure per outcome model showing all parameters (beta[1:P] + sigma).
#
#  Output: 12 individual PNGs (3 types x 4 outcomes)
#         + 3 combined PNGs (1 per diagnostic type, 4 panels each)
#
#  Prereq: bayes_results from Bayesian.Rmd
###############################################################################

library(bayesplot)
library(posterior)
library(patchwork)
library(ggplot2)

# Outcome labels for panel titles
outcome_labels <- c(
  "Change in Log10 Viral Load"  = "Delta Log10 Viral Load",
  "Change in CD4+ T Cell Count" = "Delta CD4+ T Cell Count",
  "Change in Physical QoL"      = "Delta Physical QoL",
  "Change in Mental QoL"        = "Delta Mental QoL"
)

# Parameter display labels (matches model matrix column order)
param_labels <- c(
  "beta[1]" = "Intercept",
  "beta[2]" = "Hard Drug Use",
  "beta[3]" = "Age",
  "beta[4]" = "Race",
  "beta[5]" = "Education",
  "beta[6]" = "BMI",
  "beta[7]" = "Smoking",
  "sigma"   = "Sigma"
)

# Build parameter name vector
param_names <- names(param_labels)


# =============================================================================
# HELPER: Extract draws and build a single diagnostic plot
# =============================================================================

get_draws <- function(nm) {
  r <- bayes_results[[nm]]
  as_draws_array(r$fit_adj$draws())
}


# =============================================================================
# FIGURE A: TRACE PLOTS (1 combined figure, 4 outcome panels)
# =============================================================================

trace_plots <- lapply(names(bayes_results), function(nm) {
  r <- bayes_results[[nm]]
  draws_array <- get_draws(nm)
  
  mcmc_trace(draws_array, pars = param_names,
             facet_args = list(ncol = 2, labeller = as_labeller(param_labels))) +
    ggtitle(outcome_labels[r$label]) +
    theme(plot.title      = element_text(size = 11, face = "bold"),
          strip.text       = element_text(size = 7),
          axis.text        = element_text(size = 6),
          legend.position  = "none")
})

# Save individual outcome trace plots
for (i in seq_along(trace_plots)) {
  nm <- names(bayes_results)[i]
  ggsave(paste0("fig_trace_", i, ".png"), trace_plots[[i]],
         width = 10, height = 10, dpi = 300)
}

# Combined 2x2 of all outcomes
fig_trace_all <- (trace_plots[[1]] + trace_plots[[2]]) /
                 (trace_plots[[3]] + trace_plots[[4]]) +
  plot_annotation(
    title = "Trace Plots: All Parameters (Adjusted Model)",
    subtitle = "4 chains, 10,000 post-warmup iterations each",
    theme = theme(plot.title    = element_text(size = 14, face = "bold"),
                  plot.subtitle = element_text(size = 10, colour = "grey40"))
  )

ggsave("fig_trace_all_outcomes.png", fig_trace_all,
       width = 18, height = 20, dpi = 300)


# =============================================================================
# FIGURE B: POSTERIOR DENSITY OVERLAY (1 combined figure, 4 outcome panels)
# =============================================================================

dens_plots <- lapply(names(bayes_results), function(nm) {
  r <- bayes_results[[nm]]
  draws_array <- get_draws(nm)
  
  mcmc_dens_overlay(draws_array, pars = param_names,
                    facet_args = list(ncol = 2, 
                                     labeller = as_labeller(param_labels),
                                     scales = "free")) +
    ggtitle(outcome_labels[r$label]) +
    theme(plot.title      = element_text(size = 11, face = "bold"),
          strip.text       = element_text(size = 7),
          axis.text        = element_text(size = 6),
          legend.position  = "none")
})

for (i in seq_along(dens_plots)) {
  ggsave(paste0("fig_density_", i, ".png"), dens_plots[[i]],
         width = 10, height = 10, dpi = 300)
}

fig_dens_all <- (dens_plots[[1]] + dens_plots[[2]]) /
                (dens_plots[[3]] + dens_plots[[4]]) +
  plot_annotation(
    title = "Posterior Density by Chain: All Parameters (Adjusted Model)",
    subtitle = "Overlapping densities indicate chain agreement",
    theme = theme(plot.title    = element_text(size = 14, face = "bold"),
                  plot.subtitle = element_text(size = 10, colour = "grey40"))
  )

ggsave("fig_density_all_outcomes.png", fig_dens_all,
       width = 18, height = 20, dpi = 300)


# =============================================================================
# FIGURE C: AUTOCORRELATION (1 combined figure, 4 outcome panels)
# =============================================================================

acf_plots <- lapply(names(bayes_results), function(nm) {
  r <- bayes_results[[nm]]
  draws_array <- get_draws(nm)
  
  mcmc_acf(draws_array, pars = param_names,
           facet_args = list(labeller = as_labeller(param_labels))) +
    ggtitle(outcome_labels[r$label]) +
    theme(plot.title      = element_text(size = 11, face = "bold"),
          strip.text       = element_text(size = 6),
          axis.text        = element_text(size = 5),
          legend.position  = "none")
})

for (i in seq_along(acf_plots)) {
  ggsave(paste0("fig_acf_", i, ".png"), acf_plots[[i]],
         width = 10, height = 10, dpi = 300)
}

fig_acf_all <- (acf_plots[[1]] + acf_plots[[2]]) /
               (acf_plots[[3]] + acf_plots[[4]]) +
  plot_annotation(
    title = "Autocorrelation: All Parameters (Adjusted Model)",
    subtitle = "Rapid decay to zero indicates efficient sampling",
    theme = theme(plot.title    = element_text(size = 14, face = "bold"),
                  plot.subtitle = element_text(size = 10, colour = "grey40"))
  )

ggsave("fig_acf_all_outcomes.png", fig_acf_all,
       width = 18, height = 20, dpi = 300)

```

```{r}
# =============================================================================
# 14.  SAVE ALL RESULTS
# =============================================================================
save(bayes_results, bayes_drug_effects, posterior_probs,
     bayes_mediation, fit_comparison,
     file = "bayes_results.RData")
```
