---
title: "Bayesian"
author: "Qingyang Du"
date: "2026-02-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###############################################################################
#                      BAYESIAN ANALYSIS SCRIPT
#
# Project:  Treatment Response to HAART and Hard Drug Use
# Prereq:   Run 01_data_cleaning_v3.R (produces hiv_cleaned.RData with dat_cc)
#
# Approach: Follows the professor's cmdstanr workflow:
#   1. Write a generic Stan linear regression model
#   2. Compile once, reuse for all outcomes
#   3. Prepare data (design matrix + priors) per outcome
#   4. Fit via MCMC (4 chains)
#   5. Summarize posteriors (mean, SD, MCSE, 95% HPDI, ESS)
#   6. Model fit statistics (WAIC, LOO-IC, DIC)
#   7. MCMC diagnostics (trace, density overlay, ACF, R-hat)
#
# Models (for each of 4 outcomes):
#   Model 1: Adjusted — Y2 ~ hard_drugs_bl + Y_bl + confounders
#   Model 2: Adjusted + Adherence — Model 1 + ADH_cat
###############################################################################


# =============================================================================
# 0.  LOAD PACKAGES AND DATA
# =============================================================================
```{r}
# install.packages("cmdstanr", repos = c('https://stan-dev.r-universe.dev',
#                                         getOption("repos")))
# library(cmdstanr)
# install_cmdstan()   # Run once to install CmdStan backend

library(cmdstanr)
library(bayesplot)   # MCMC diagnostic plots
library(posterior)   # summarizing posterior draws
library(bayestestR)  # highest density posterior intervals (hdi)
library(mcmcse)      # Monte Carlo standard errors
library(loo)         # WAIC and LOO-IC
library(dplyr)
library(tibble)

load("hiv_cleaned.RData")   # contains dat_cc




# =============================================================================
# 1.  WRITE THE STAN MODEL (reusable for all outcomes)
# =============================================================================
# Same structure as the professor's example: normal priors on beta,
# half-normal prior on sigma, generated quantities for log-likelihood.

stan_code <- "data {
  int<lower=0> N;                  // number of observations
  int<lower=0> P;                  // number of predictors including intercept
  matrix[N, P] X;                  // design matrix (first column = intercept)
  vector[N] y;                     // outcome

  vector[P] prior_mean;            // prior means for each beta
  vector<lower=0>[P] prior_sd;     // prior SDs for each beta

  real<lower=0> sigma_prior_sd;    // SD for half-normal prior on sigma
}

parameters {
  vector[P] beta;                  // regression coefficients
  real<lower=0> sigma;             // residual SD
}

model {
  // Priors for regression coefficients
  beta ~ normal(prior_mean, prior_sd);

  // Half-normal prior for sigma
  sigma ~ normal(0, sigma_prior_sd);

  // Likelihood
  y ~ normal(X * beta, sigma);
}

generated quantities {
  // Log likelihood for each observation (for WAIC / LOO-IC)
  vector[N] log_lik;
  for (n in 1:N) {
    log_lik[n] = normal_lpdf(y[n] | X[n] * beta, sigma);
  }
}
"

stan_file <- write_stan_file(stan_code, dir = ".", basename = "linear_regression")


# =============================================================================
# 2.  COMPILE THE STAN MODEL
# =============================================================================

mod <- cmdstan_model(stan_file)


# =============================================================================
# 3.  DEFINE OUTCOMES AND PRIOR SPECIFICATIONS
# =============================================================================

# Outcome definitions — same structure as frequentist script
outcomes <- list(
  list(name  = "delta_log_VLOAD",
       y     = "delta_log_VLOAD",
       label = "Change in Log10 Viral Load"),
  list(name  = "delta_LEU3N",
       y     = "delta_LEU3N",
       label = "Change in CD4+ T Cell Count"),
  list(name  = "delta_AGG_PHYS",
       y     = "delta_AGG_PHYS",
       label = "Change in Physical QoL"),
  list(name  = "delta_AGG_MENT",
       y     = "delta_AGG_MENT",
       label = "Change in Mental QoL")
)

# Confounders (same for all outcomes)
confounders <- c("age_bl", "RACE_bin_bl", "EDUC_bin_bl", "BMI_bl", "SMOKE_bin_bl")


# =============================================================================
# 4.  HELPER FUNCTION: PREPARE DATA AND SET PRIORS
# =============================================================================

prepare_stan_data <- function(dat, outcome_info, include_adherence = FALSE) {

  # Build the model formula
  rhs <- c("hard_drugs_bl", confounders)
  if (include_adherence) rhs <- c(rhs, "ADH_cat")
  f <- as.formula(paste(outcome_info$y, "~", paste(rhs, collapse = " + ")))

  # Design matrix
  X <- model.matrix(f, data = dat)
  y <- dat[[outcome_info$y]]

  N <- nrow(X)
  P <- ncol(X)

  # --- Prior specification ---
  # Weakly informative priors scaled to the outcome:
  #   Intercept: Normal(mean(y), 2 * sd(y))  — centered on the data mean
  #   All other coefficients: Normal(0, sd(y)) — centered at no effect,
  #     scaled to the outcome's SD so they are weakly informative
  #   Sigma: half-Normal(0, 2 * sd(y))

  y_sd <- sd(y)
  y_mean <- mean(y)

  m <- c(y_mean, rep(0, P - 1))       # prior means
  s <- c(2 * y_sd, rep(y_sd, P - 1))  # prior SDs
  sigma_sd <- 2 * y_sd                 # sigma prior SD

  data_list <- list(
    N = N,
    P = P,
    X = X,
    y = y,
    prior_mean = m,
    prior_sd = s,
    sigma_prior_sd = sigma_sd
  )

  list(data_list = data_list, X = X, y = y, P = P, N = N,
       col_names = colnames(X))
}


# =============================================================================
# 5.  HELPER FUNCTION: SUMMARIZE POSTERIOR
# =============================================================================

summarize_posterior <- function(fit, col_names) {
  draws_mat <- as_draws_matrix(fit$draws())
  params <- colnames(draws_mat)
  params <- params[!grepl("lp__|log_lik", params)]

  summary_table <- lapply(params, function(p) {
    vals <- as.numeric(draws_mat[, p])

    # Monte Carlo standard error
    mcse_val <- mcmcse::mcse(vals)$se

    # Effective sample size
    ess_val <- ess_bulk(vals)

    # 95% Highest Posterior Density Interval
    hpd <- hdi(vals, ci = 0.95)

    tibble(
      Parameter     = p,
      Estimate      = mean(vals),
      MCSE          = mcse_val,
      Posterior_SD  = sd(vals),
      HPDI_2.5      = hpd$CI_low,
      HPDI_97.5     = hpd$CI_high,
      ESS           = ess_val
    )
  }) %>% bind_rows()

  # Map beta indices to variable names
  beta_idx <- grep("^beta\\[", summary_table$Parameter)
  if (length(beta_idx) == length(col_names)) {
    summary_table$Variable <- NA_character_
    summary_table$Variable[beta_idx] <- col_names
    summary_table$Variable[summary_table$Parameter == "sigma"] <- "sigma"
    summary_table <- summary_table %>%
      select(Parameter, Variable, everything())
  }

  # Check MCSE < 6% of posterior SD
  summary_table <- summary_table %>%
    mutate(MCSE_pct_SD = round(100 * MCSE / Posterior_SD, 2),
           MCSE_OK     = MCSE_pct_SD < 6)

  summary_table
}


# =============================================================================
# 6.  HELPER FUNCTION: MODEL FIT STATISTICS
# =============================================================================

compute_fit_stats <- function(fit, X, y) {
  draws_mat <- as_draws_matrix(fit$draws())
  loglik_mat <- as_draws_matrix(fit$draws("log_lik"))

  # LOO-IC
  loo_res <- loo(loglik_mat)

  # WAIC
  waic_res <- waic(loglik_mat)

  # DIC (computed by hand, following professor's approach)
  D_theta <- -2 * rowSums(loglik_mat)
  mean_D <- mean(D_theta)

  beta_mean <- colMeans(draws_mat[, grep("^beta\\[", colnames(draws_mat))])
  sigma_mean <- mean(draws_mat[, "sigma"])
  D_bar_theta <- -2 * sum(dnorm(y, mean = X %*% beta_mean,
                                 sd = sigma_mean, log = TRUE))
  DIC <- 2 * mean_D - D_bar_theta

  list(loo = loo_res, waic = waic_res, DIC = DIC)
}


# =============================================================================
# 7.  HELPER FUNCTION: MCMC DIAGNOSTICS
# =============================================================================

run_diagnostics <- function(fit, col_names, outcome_label, model_label) {

  draws_array <- as_draws_array(fit$draws())

  # Build parameter names for diagnostic plots
  P <- length(col_names)
  param_names <- c(paste0("beta[", 1:P, "]"), "sigma")

  # Trace plots
  p_trace <- mcmc_trace(draws_array, pars = param_names) +
    ggtitle(paste(outcome_label, "-", model_label, ": Trace Plots"))
  print(p_trace)

  # Posterior density overlay across chains
  p_dens <- mcmc_dens_overlay(draws_array, pars = param_names) +
    ggtitle(paste(outcome_label, "-", model_label, ": Posterior Densities"))
  print(p_dens)

  # Autocorrelation
  p_acf <- mcmc_acf(draws_array, pars = param_names) +
    ggtitle(paste(outcome_label, "-", model_label, ": Autocorrelation"))
  print(p_acf)

  # R-hat and ESS from cmdstanr
  rhat_summary <- fit$summary(variables = param_names)
  print(rhat_summary)

  # CmdStan diagnostics (divergences, treedepth, energy)
  fit$cmdstan_diagnose()
}

```

```{r}
# =============================================================================
# 8.  FIT ALL MODELS
# =============================================================================

bayes_results <- list()

for (o in outcomes) {

  message("\n============================================================")
  message("FITTING: ", o$label)
  message("============================================================\n")

  # --- Model 1: Adjusted (no adherence) ---
  prep1 <- prepare_stan_data(dat_cc, o, include_adherence = FALSE)

  fit1 <- mod$sample(
    data = prep1$data_list,
    chains = 4,
    iter_warmup = 2500,
    iter_sampling = 10000,
    seed = 42,
    parallel_chains = 4
  )

  summ1 <- summarize_posterior(fit1, prep1$col_names)
  fits1 <- compute_fit_stats(fit1, prep1$X, prep1$y)

  message("\n--- Adjusted Model Summary ---")
  print(summ1)
  message("DIC: ", round(fits1$DIC, 2))

  # --- Model 2: Adjusted + Adherence ---
  prep2 <- prepare_stan_data(dat_cc, o, include_adherence = TRUE)

  fit2 <- mod$sample(
    data = prep2$data_list,
    chains = 4,
    iter_warmup = 2500,
    iter_sampling = 10000,
    seed = 42,
    parallel_chains = 4
  )

  summ2 <- summarize_posterior(fit2, prep2$col_names)
  fits2 <- compute_fit_stats(fit2, prep2$X, prep2$y)

  message("\n--- Adjusted + Adherence Model Summary ---")
  print(summ2)
  message("DIC: ", round(fits2$DIC, 2))

  # Store results
  bayes_results[[o$name]] <- list(
    label        = o$label,
    # Adjusted model
    fit_adj      = fit1,
    summary_adj  = summ1,
    fitstats_adj = fits1,
    prep_adj     = prep1,
    # Adjusted + adherence
    fit_adj_adh      = fit2,
    summary_adj_adh  = summ2,
    fitstats_adj_adh = fits2,
    prep_adj_adh     = prep2
  )
}


# =============================================================================
# 9.  EXTRACT HARD DRUG USE EFFECT ACROSS ALL MODELS
# =============================================================================

extract_drug_bayes <- function(summary_tbl, model_label, outcome_label) {
  summary_tbl %>%
    filter(Variable == "hard_drugs_blYes") %>%
    mutate(Outcome = outcome_label, Model = model_label) %>%
    select(Outcome, Model, Estimate, Posterior_SD, HPDI_2.5, HPDI_97.5,
           ESS, MCSE, MCSE_pct_SD, MCSE_OK)
}

bayes_drug_effects <- map_dfr(names(bayes_results), function(nm) {
  r <- bayes_results[[nm]]
  bind_rows(
    extract_drug_bayes(r$summary_adj,     "Adjusted",             r$label),
    extract_drug_bayes(r$summary_adj_adh, "Adjusted + Adherence", r$label)
  )
})

bayes_drug_effects <- bayes_drug_effects %>%
  mutate(across(c(Estimate, Posterior_SD, HPDI_2.5, HPDI_97.5, MCSE),
                ~ round(., 4)),
         ESS = round(ESS, 0))

bayes_drug_effects


# =============================================================================
# 10.  POSTERIOR PROBABILITY OF HYPOTHESIZED DIRECTION
# =============================================================================
# H1: hard drug users have WORSE outcomes at year 2:
#   log_VLOAD: higher  → P(beta > 0)
#   LEU3N:     lower   → P(beta < 0)
#   AGG_PHYS:  lower   → P(beta < 0)
#   AGG_MENT:  lower   → P(beta < 0)

hypothesis_direction <- c(
  "log_VLOAD" = "greater",    # beta > 0 means higher viral load
  "LEU3N"     = "less",       # beta < 0 means lower CD4
  "AGG_PHYS"  = "less",       # beta < 0 means lower physical QoL
  "AGG_MENT"  = "less"        # beta < 0 means lower mental QoL
)

posterior_probs <- map_dfr(names(bayes_results), function(nm) {
  r <- bayes_results[[nm]]
  direction <- hypothesis_direction[[nm]]

  # Extract hard_drugs_bl draws from adjusted model
  draws_mat <- as_draws_matrix(r$fit_adj$draws())
  drug_col <- which(r$prep_adj$col_names == "hard_drugs_blYes")
  drug_draws <- as.numeric(draws_mat[, paste0("beta[", drug_col, "]")])

  if (direction == "greater") {
    prob <- mean(drug_draws > 0)
  } else {
    prob <- mean(drug_draws < 0)
  }

  tibble(
    Outcome    = r$label,
    Direction  = ifelse(direction == "greater",
                        "P(beta > 0)", "P(beta < 0)"),
    Probability = round(prob, 4)
  )
})

posterior_probs


# =============================================================================
# 11.  MEDIATION COMPARISON (BAYESIAN)
# =============================================================================

bayes_mediation <- bayes_drug_effects %>%
  select(Outcome, Model, Estimate) %>%
  pivot_wider(names_from = Model, values_from = Estimate,
              names_prefix = "est_") %>%
  rename(est_adj     = `est_Adjusted`,
         est_adj_adh = `est_Adjusted + Adherence`) %>%
  mutate(
    change     = est_adj_adh - est_adj,
    pct_change = round(100 * change / est_adj, 1)
  )

bayes_mediation


# =============================================================================
# 12.  MCMC DIAGNOSTICS — ALL MODELS
# =============================================================================

for (nm in names(bayes_results)) {
  r <- bayes_results[[nm]]

  message("\n--- Diagnostics: ", r$label, " (Adjusted) ---")
  run_diagnostics(r$fit_adj, r$prep_adj$col_names,
                  r$label, "Adjusted")

  message("\n--- Diagnostics: ", r$label, " (Adjusted + Adherence) ---")
  run_diagnostics(r$fit_adj_adh, r$prep_adj_adh$col_names,
                  r$label, "Adjusted + Adherence")
}


# =============================================================================
# 13.  MODEL FIT COMPARISON TABLE
# =============================================================================

fit_comparison <- map_dfr(names(bayes_results), function(nm) {
  r <- bayes_results[[nm]]
  tibble(
    Outcome         = r$label,
    # Adjusted model
    DIC_adj         = round(r$fitstats_adj$DIC, 2),
    WAIC_adj        = round(r$fitstats_adj$waic$estimates["waic", "Estimate"], 2),
    LOOIC_adj       = round(r$fitstats_adj$loo$estimates["looic", "Estimate"], 2),
    # Adjusted + adherence
    DIC_adj_adh     = round(r$fitstats_adj_adh$DIC, 2),
    WAIC_adj_adh    = round(r$fitstats_adj_adh$waic$estimates["waic", "Estimate"], 2),
    LOOIC_adj_adh   = round(r$fitstats_adj_adh$loo$estimates["looic", "Estimate"], 2)
  )
})

fit_comparison


# =============================================================================
# 14.  SAVE ALL RESULTS
# =============================================================================

save(bayes_results, bayes_drug_effects, posterior_probs,
     bayes_mediation, fit_comparison,
     file = "bayes_results.RData")

```